---
title: "ATLANTIS_exacerbations"
author: "Tatiana Karp, Rui Mercalo"
date: "2025-07-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = "~/Work/RP2/ATLANTIS")
library(dplyr)
library(edgeR)
library(ggplot2)
library(GSVA)
library(pROC)
library(biomaRt)
library(ggrepel)
library(stringr)

```

## 

## Prepare the metadata

```{r metadata, echo = FALSE}
atlantis_exacerbations <- read.csv(file = './Exacerbations/ATLANTIS_master_table_exacerbation.csv')
atlantis_exacerbations <- atlantis_exacerbations%>%
  dplyr::select(-X)%>%
  filter (QC_check == 'YES')%>%
  mutate(gender = as.factor(gender),
         smoking.status = as.factor(smoking.status),
         exacerbation = as.factor (exacerbation),
         time_to_exacerbation = as.numeric(time_to_exacerbation),
         follow_up = as.numeric(follow_up),
         Ex_freq = as.factor(if_else(NUM_EX_D>1, 'freq', 'not_freq')))

# filter out samples with follow-up less than 250 days
atlantis_exacerbations_decreased <- atlantis_exacerbations%>%
  filter(follow_up >= 250)

```

## Add expression data and perform differentiall gene expression analysis 

```{r DGE, echo=FALSE}
# expression data
expression.data <- read.csv('./Umi_dedup/20201107_ATLANTIS_raw_readcount_dedup_FINAL.csv', header =TRUE)%>%
  tibble::column_to_rownames("Gene")%>%
  dplyr::select(c(atlantis_exacerbations_decreased$GenomeScan_ID))%>%
  as.matrix()

all(colnames(expression.data) == atlantis_exacerbations_decreased$GenomeScan_ID)

# DE yes/no exacerbations
design <- model.matrix(~0 + exacerbation + age + gender + smoking.status, data = atlantis_exacerbations_decreased)

DGEL_exac <- edgeR::DGEList(expression.data)
keep <- edgeR::filterByExpr(DGEL_exac, design) 
DGEL_exac <- DGEL_exac[keep, , keep.lib.sizes=FALSE]
DGEL_exac <- edgeR::calcNormFactors(DGEL_exac, method = "TMM")

TMM_expr <- cpm(DGEL_exac, normalized.lib.sizes = TRUE, log = TRUE) # save cpm normalized for later
#write.csv(TMM_expr, "./Exacerbations/TMM_expr_atl.csv")


# Fit the model
DGEL_exac <- edgeR::estimateDisp(DGEL_exac, design)
fit_exac <- edgeR::glmQLFit(DGEL_exac, design, legacy = T)
contrasts <- limma::makeContrasts(EX_binary = exacerbationTRUE - exacerbationFALSE,
                                  levels = design)
qlf_exac   <- edgeR::glmQLFTest(fit_exac, contrast = contrasts[,"EX_binary"])
summary(decideTests(qlf_exac))

de.results.exac <- edgeR::topTags(
  qlf_exac,
  n=nrow(DGEL_exac))$table %>%
  tibble::rownames_to_column("Gene") 

# add gene names
ensembl38 <- useEnsembl(biomart = "genes",
                        dataset = "hsapiens_gene_ensembl",
                        version = 101)

dataprobes <- getBM(filters = "ensembl_gene_id", attributes = c(
  "ensembl_gene_id",
  "external_gene_name"),
  values = as.character(de.results.exac$Gene),
  mart = ensembl38
)

de.results.exac <- de.results.exac %>% 
  dplyr::left_join(
    y = dataprobes,
    by = c("Gene" = "ensembl_gene_id")) 

# write.csv(de.results.exac %>%
#             mutate(across(where(is.numeric), ~ signif(.x, digits = 3))),
#           "./Exacerbations/DE.genes.exac.yes.no.303.csv")
```
## Volcano plot 

```{r volcano, echo=FALSE}
# de.results.exac <- de.results.exac %>%
#   mutate(names_to_plot = if_else(str_starts(external_gene_name, "AC"), Gene, external_gene_name))

volcano <- ggplot(data = de.results.exac, aes(x=logFC, y=-log10(FDR))) +
  geom_point(aes(color= ifelse((FDR < 0.05)&(logFC>0),'firebrick', ifelse((FDR < 0.05)&(logFC<0),'blue','gray')))) + 
  theme_classic()+
  geom_text_repel(aes(label=ifelse(((FDR < 0.05 ) & (abs(logFC)>1)) | (FDR < 0.000001),
                                   (ifelse((!is.na(external_gene_name)), external_gene_name, '')), ''),
                      lineheight=0.5, hjust= 0.5, vjust= 0.4),
                  max.overlaps =50,
                  direction = 'both',
                  alpha = 0.6,
                  box.padding=0.3,
                  point.padding=0.5)+
  xlab (expression (log[2]~fold~change))+
  ylab (expression(-log[10]~FDR))+
  scale_color_identity(name = '', breaks= c('firebrick', 'blue','gray') ,
                       labels = c('higher in exacerbators',
                                  'lower in exacerbators', 
                                  'not significant'), guide = "legend")+
  xlim(-2.5, 2.5) + 
  theme(axis.text = element_text(size=15),
        axis.title = element_text(size=15, face="plain"),
        legend.position = "bottom",
        panel.grid.major =  element_line(colour = "grey70", size = 0.1), panel.grid.minor = element_blank(),
        panel.background = element_blank(), axis.line = element_line(colour = "black", linewidth =0.6),
        legend.text = element_text(size=12),
        legend.key = element_rect(fill = "white"),
        plot.title = element_text(hjust = 0.5, size = 15))

saveRDS(volcano, "./Exacerbations/Exacerbations_DE_volcano.rds")

png("./Exacerbations/Exacerbations_DE_volcano.png",
    width = 1100, 
    height = 1200,
    res = 150)
print(volcano)
dev.off()
```

## Calculate GSVA score for DE genes

```{r gsva, echo=FALSE}
# define gene set - significant DE genes: 
geneSet<- list(exacerbation_genes = de.results.exac[de.results.exac$FDR<0.05, 'Gene'])

# define GSVA parameters: 
# maxDiff - FALSE - to include both up and down genes
# kcdf -Gaussian - for logCPM normalized genes
gsvapar <- gsvaParam(as.matrix(TMM_expr), geneSet, maxDiff = FALSE, kcdf = "Gaussian")   
gsva_score <- gsva(gsvapar, verbose = TRUE) %>%
  t() %>%
  as.data.frame() %>%
  tibble::rownames_to_column('Sample')

## add gsva to clinical table 
atlantis_exacerbations_decreased <- atlantis_exacerbations_decreased %>%
  left_join(gsva_score, by = c("GenomeScan_ID" = "Sample")) %>%
  rename("gsva_score_exacerbation_genes" = "exacerbation_genes")

```

## Prepare clinical data for exacerbation prediction

```{r prepare_clinical_model, echo=FALSE}

clinical_table <-  read.csv('./atlantis_patient_data.csv', header = TRUE, na.strings = c("","NA")) %>%
  dplyr:: select(c('PT','GINA','LABEOSV', 'B_FEV1PNVG','acq6_score','NUM_EX'))
clinical_table <- clinical_table[!duplicated(clinical_table$PT), ] %>% #choose data only from first visit
  mutate (GINA = as.factor(GINA))


model_table <- atlantis_exacerbations_decreased %>%
  left_join(clinical_table, by = c('PT'='PT'))  %>%
  dplyr::select(c(gender, smoking.status, GenomeScan_ID, GINA, LABEOSV, B_FEV1PNVG, Ex_freq,
                  exacerbation, follow_up, NUM_EX_D, NUM_EX, acq6_score, gsva_score_exacerbation_genes)) %>%
  mutate(ex_last_y = as.factor(if_else(NUM_EX>0, 'YES', 'NO'))) %>%
  filter(!is.na(LABEOSV)) %>%  ## 2 removed with 0.00
  filter(!is.na(B_FEV1PNVG)) %>% ## 3 removed with NAs
  mutate(GINA_mild_severe = if_else(GINA %in% c('1','2','3'), 'mild', 'severe'))%>%
  mutate(smoking = if_else (smoking.status %in% c('Non.smoker'), 'never', 'ex_cur_smoker'))

table(model_table$exacerbation)

```


## Predict exacerbations only using clinical variables (clinical model)

```{r clinical_model, echo=FALSE}

model_clinical = glm(exacerbation ~ gender + smoking + acq6_score +  B_FEV1PNVG + ex_last_y + LABEOSV + GINA_mild_severe, data = model_table, family = binomial )
summary(model_clinical)

model_table_clinical <- model_table%>%
  mutate(mod_clinical_l_pred = round(stats::predict(model_clinical, type = 'response')))

actual <- (if_else(model_table_clinical$exacerbation == TRUE, 1,0))
predicted_clinical <- model_table_clinical$mod_clinical_l_pred
outcomes_clinical <- table(predicted_clinical, actual)

# Model Accuracy
prob_clinical = stats::predict(model_clinical, type = "response")
roc_clinical = roc(model_table_clinical$exacerbation ~ prob_clinical, plot = TRUE, print.auc = TRUE)

mean(predicted_clinical == actual)
auc(roc_clinical)
outcomes_clinical


# Extract confusion matrix values
TN_clinical <- outcomes_clinical[1,1]
FP_clinical <- outcomes_clinical[2,1]
FN_clinical <- outcomes_clinical[1,2]
TP_clinical <- outcomes_clinical[2,2]

# Sensitivity and Specificity
sensitivity_clinical <- TP_clinical / (TP_clinical + FN_clinical)
specificity_clinical <- TN_clinical / (TN_clinical + FP_clinical)

# Balanced Accuracy -arithmetic mean of sensitivity and specificity
balanced_accuracy_clinical <- (sensitivity_clinical + specificity_clinical) / 2
balanced_accuracy_clinical # 0.6203569


```


```{r gsva_model, echo=FALSE}

model_GSVA = glm(exacerbation ~ gender + smoking + acq6_score +  B_FEV1PNVG + ex_last_y + LABEOSV + GINA_mild_severe + gsva_score_exacerbation_genes, data = model_table, family = binomial)
summary(model_GSVA)

model_table_GSVA <- model_table%>%
  mutate(mod_GSVA_l_pred = round(stats::predict(model_GSVA, type = 'response')))

actual <- (if_else(model_table_GSVA$exacerbation == TRUE, 1,0))
predicted_GSVA <- model_table_GSVA$mod_GSVA_l_pred
outcomes_GSVA <- table(predicted_GSVA, actual)

# Model Accuracy
prob_GSVA = stats::predict(model_GSVA, type = "response")
roc_GSVA = roc(model_table_GSVA$exacerbation ~ prob_GSVA, plot = TRUE, print.auc = TRUE)


mean(predicted_GSVA == actual)
auc(roc_GSVA)
outcomes_GSVA


# Extract confusion matrix values
TN_GSVA <- outcomes_GSVA[1,1]
FP_GSVA <- outcomes_GSVA[2,1]
FN_GSVA <- outcomes_GSVA[1,2]
TP_GSVA <- outcomes_GSVA[2,2]

# Sensitivity and Specificity
sensitivity_GSVA <- TP_GSVA / (TP_GSVA + FN_GSVA)
specificity_GSVA <- TN_GSVA / (TN_GSVA + FP_GSVA)

# Balanced Accuracy
balanced_accuracy_GSVA <- (sensitivity_GSVA + specificity_GSVA) / 2
balanced_accuracy_GSVA # 0.6676696

```

Compare clinical vs GSVA_model 

```{r compare_model, echo=FALSE}

# Compare the Optimal vs Clinical models in ATLANTIS 

ATLANTIS_Clinical_vs_GSVA <- anova(model_clinical, model_GSVA, test = "Chisq") ## 
ATLANTIS_Clinical_vs_GSVA # p-value = 0.001366 

```

Plot ROC for clinical and GSVA models

```{r roc, echo=FALSE}

# ROC objects
ATLANTIS_roc_clinical <- roc(model_table$exacerbation, prob_clinical)
ATLANTIS_roc_GSVA <- roc(model_table$exacerbation, prob_GSVA)

# Combine into data frame for ggplot
ATLANTIS_roc <- data.frame(
  fpr = c(1 - ATLANTIS_roc_clinical$specificities, 1 - ATLANTIS_roc_GSVA$specificities),
  tpr = c(ATLANTIS_roc_clinical$sensitivities, ATLANTIS_roc_GSVA$sensitivities),
  model = rep(c("Clinical", "GSVA"), each = length(ATLANTIS_roc_clinical$specificities))
)

# Plot
ggplot(ATLANTIS_roc, aes(x = fpr, y = tpr, color = model)) +
  geom_line(linewidth = 1.2) +
  geom_abline(linetype = "dashed", color = "gray") +
  labs(
    title = "ROC Curve: Clinical vs Clinical + GSVA (ATLANTIS)",
    x = "False Positive Rate (1 - Specificity)",
    y = "True Positive Rate (Sensitivity)"
  ) +
  theme_minimal() +
  scale_color_manual(values = c("blue", "red"))

```
Stepwise model building approach, incrementally adding covariates to assess their contribution to model performance (AUC, accuracy).
Incremental model building

```{r incremental_model echo=FALSE}

# Function to fit model and compute metrics
fit_model <- function(predictors, data, model_id) {
  formula_str <- paste("exacerbation ~", paste(predictors, collapse = " + "))
  model <- glm(as.formula(formula_str), data = data, family = binomial)

  # Predicted probabilities and classification
  prob <- stats::predict(model, type = "response")
  pred_class <- round(prob)
  actual <- if_else(data$exacerbation == TRUE, 1, 0)
  #confusion matrix
  outcomes <- table(pred_class, actual)
  
  # Default values
  sensitivity <- 0
  specificity <- 0

  if (nrow(outcomes) == 2 && ncol(outcomes) == 2) {
    TP <- outcomes["1", "1"]
    FN <- outcomes["0", "1"]
    TN <- outcomes["0", "0"]
    FP <- outcomes["1", "0"]

    if ((TP + FN) > 0) sensitivity <- TP / (TP + FN)
    if ((TN + FP) > 0) specificity <- TN / (TN + FP)
  } else {
    message("Warning: Confusion matrix does not have both classes. Setting sensitivity and specificity to 0.")
  }
  
  # ROC & AUC
  roc_obj <- roc(data$exacerbation ~ prob, plot = TRUE, print.auc = TRUE, main = paste("Model", model_id))
  auc_value <- auc(roc_obj)
  acc <- mean(pred_class == actual)

  # Return results
  return(list(
    model_id = model_id,
    model = model,
    formula = formula_str,
    auc = auc_value,
    accuracy = acc,
    balanced_accuracy = (sensitivity + specificity) / 2,
    sensitivity = sensitivity,
    confusion = outcomes,
    roc = roc_obj
  ))
}


# List of predictors to add step-by-step
predictor_sets <- list(
  c("gender", "smoking"),
  c("gender", "smoking", "acq6_score"),
  c("gender", "smoking", "acq6_score", "B_FEV1PNVG"),
  c("gender", "smoking", "acq6_score", "B_FEV1PNVG", "ex_last_y"),
  c("gender", "smoking", "acq6_score", "B_FEV1PNVG", "ex_last_y", "LABEOSV"),
  c("gender", "smoking", "acq6_score", "B_FEV1PNVG", "ex_last_y", "LABEOSV", "GINA_mild_severe"),
  c("gender", "smoking", "acq6_score", "B_FEV1PNVG", "ex_last_y", "LABEOSV", "GINA_mild_severe", "gsva_score_exacerbation_genes")
)

# Run models
model_results <- lapply(seq_along(predictor_sets), function(i) {
  fit_model(predictors = predictor_sets[[i]], data = model_table, model_id = paste0("Model_", i))
})

# Create summary table

model_summary <- model_results %>%
  lapply(function(x) {
    tibble(
      Model = x$model_id,
      Formula = x$formula,
      Accuracy = round(x$accuracy, 3),
      AUC = round(as.numeric(x$auc), 3),
      Sensitivity = round(x$sensitivity, 3),
      Balanced_accuracy = round(x$balanced_accuracy, 3)
    )
  }) %>%
  bind_rows()

```

```{r}

model_perf <- data.frame(
  Model = c(
    "Sex + Smoking",
    "+ ACQ-6",
    "+ FEV1 % pred",
    "+ Exacerbation history",
    "+ Blood eosinophils",
    "+ GINA",
    "+ GSVA"
  ),
  Sensitivity = (model_summary$Sensitivity*100)
)

# Factor to keep the order
model_perf$Model <- factor(model_perf$Model, levels = model_perf$Model)

# Custom labels with expression (subscript for FEV[1])
custom_labels <- c(
  "Sex + Smoking",
  "+ ACQ-6",
  expression("+ FEV"[1]*" % pred"),
  "+ Exacerbation history",
  "+ Blood eosinophils",
  "+ GINA",
  "+ GSVA"
)


ggplot(model_perf, aes(x = Model, y = Sensitivity)) +
  geom_bar(stat = "identity", fill = "#2c7fb8") +
  geom_text(aes(label = paste0(Sensitivity, "%")), vjust = -0.5, size = 4) +
  labs(
    title = "Change in sensitivity (True positive rate) with each additional variable",
    x = "Model (Cumulative Addition of Predictors)",
    y = "Sensitivity (%)"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) +
  scale_x_discrete(labels = custom_labels) +
  ylim(0, 45)

```

