---
title: "Replication_exacerbations_in_alliance - One data per participant"
author: "Tatiana Karp, Rui Mercalo"
date: "2025-10-06"
output: html_document
---

```{r setup, include=FALSE}
library(readxl)
library(dplyr)
library(pROC)
library(here)

source(here("config.R"))
knitr::opts_knit$set(root.dir = ROOT_DIR)

# Source the cross-validation helper functions
source(here("R/cv_functions.R"))

```

## Load Alliance data
### One datapoint per participant 

```{r metadata, echo = FALSE}
sample_list <- read_excel(PATHS$alliance_meta) 
alliance_GSVA <- read.csv(PATHS$alliance_scores)
# matching ids: pseudo.t_timepoint_1
```

## Prepare clinical data for exacerbation prediction

```{r prepare_clinical_model, echo=FALSE}
model_table <- sample_list %>%
  left_join(alliance_GSVA, by = c("pseudo.t_timepoint_1" = "X")) %>%
  dplyr::select(c(gender, smoking, GINA_mild_severe, LABEOSV, B_FEV1PNVG, ex_last_y,
                  exacerbation, acq6_score, exacerbation_genes, TNFAIP6, ENSG00000287937)) %>%
  filter(!is.na(LABEOSV)) %>%  ## 1 removed 
  filter(!is.na(B_FEV1PNVG)) %>% ## 1 removed 
  filter(!is.na(smoking)) %>% # 1 removed
  filter(!(smoking == "NA")) %>% #one sample had character "NA"
  filter(!is.na(ex_last_y)) %>%
  mutate(exacerbation = if_else(exacerbation == "Yes", 1, 0)) %>%
  mutate(exacerbation_f = factor(if_else(exacerbation == 1, "Yes", "No"), levels = c("Yes", "No")))


table(model_table$exacerbation)
```

## Predict exacerbations only using clinical variables (clinical model)
## Logistic regression: in-sample model fit

```{r clinical_model, echo=FALSE}

model_clinical = glm(exacerbation ~ gender + smoking + acq6_score +  B_FEV1PNVG + ex_last_y + LABEOSV + GINA_mild_severe, data = model_table, family = binomial )
summary(model_clinical)

coef_table <- summary(model_clinical)$coefficients %>%
  as.data.frame() %>%
  tibble::rownames_to_column("term") %>%
  as_tibble() %>%
  mutate(across(where(is.numeric), ~ round(.x, 3)))
                              
write.csv(coef_table, OUTPUTS$alliance_model_clinical)

# model_table_clinical <- model_table%>%
#   mutate(mod_clinical_l_pred = round(stats::predict(model_clinical, type = 'response')))
# 
# actual <- (if_else(model_table_clinical$exacerbation == TRUE, 1,0))
# predicted_clinical <- model_table_clinical$mod_clinical_l_pred
# outcomes_clinical <- table(predicted_clinical, actual)
# 
# # Model Accuracy
# prob_clinical = stats::predict(model_clinical, type = "response")
# roc_clinical = roc(model_table_clinical$exacerbation ~ prob_clinical, plot = TRUE, print.auc = TRUE)
# 
# mean(predicted_clinical == actual) # accuracy
# auc(roc_clinical)
# outcomes_clinical
# 
# # Extract confusion matrix values
# TN_clinical <- outcomes_clinical[1,1]
# FP_clinical <- outcomes_clinical[2,1]
# FN_clinical <- outcomes_clinical[1,2]
# TP_clinical <- outcomes_clinical[2,2]
# 
# # Sensitivity and Specificity
# sensitivity_clinical <- TP_clinical / (TP_clinical + FN_clinical)
# specificity_clinical <- TN_clinical / (TN_clinical + FP_clinical)
# 
# # Balanced Accuracy -arithmetic mean of sensitivity and specificity
# balanced_accuracy_clinical <- (sensitivity_clinical + specificity_clinical) / 2
# balanced_accuracy_clinical # 0.5652443

```

## Logistic regression with GSVA score: in-sample model fit

```{r gsva_model, echo=FALSE}

model_GSVA = glm(exacerbation ~ gender + smoking + acq6_score +  B_FEV1PNVG + ex_last_y + LABEOSV + GINA_mild_severe + exacerbation_genes, data = model_table, family = binomial)
summary(model_GSVA)

coef_table <- summary(model_GSVA)$coefficients %>%
  as.data.frame() %>%
  tibble::rownames_to_column("term") %>%
  as_tibble() %>%
  mutate(across(where(is.numeric), ~ round(.x, 3)))
                              
write.csv(coef_table, OUTPUTS$alliance_model_score)
# 
# model_table_GSVA <- model_table%>%
#   mutate(mod_GSVA_l_pred = round(stats::predict(model_GSVA, type = 'response')))
# 
# actual <- (if_else(model_table_GSVA$exacerbation == TRUE, 1,0))
# predicted_GSVA <- model_table_GSVA$mod_GSVA_l_pred
# outcomes_GSVA <- table(predicted_GSVA, actual)
# 
# # Model Accuracy
# prob_GSVA = stats::predict(model_GSVA, type = "response")
# roc_GSVA = roc(model_table_GSVA$exacerbation ~ prob_GSVA, plot = TRUE, print.auc = TRUE)
# 
# 
# mean(predicted_GSVA == actual)
# auc(roc_GSVA)
# outcomes_GSVA
# 
# 
# # Extract confusion matrix values
# TN_GSVA <- outcomes_GSVA[1,1]
# FP_GSVA <- outcomes_GSVA[2,1]
# FN_GSVA <- outcomes_GSVA[1,2]
# TP_GSVA <- outcomes_GSVA[2,2]
# 
# # Sensitivity and Specificity
# sensitivity_GSVA <- TP_GSVA / (TP_GSVA + FN_GSVA)
# specificity_GSVA <- TN_GSVA / (TN_GSVA + FP_GSVA)
# 
# # Balanced Accuracy
# balanced_accuracy_GSVA <- (sensitivity_GSVA + specificity_GSVA) / 2
# balanced_accuracy_GSVA #  0.6258503
```

Compare clinical vs GSVA_model 

```{r compare_model, echo=FALSE}

# Compare the Optimal vs Clinical models in ATLANTIS 

Alliance_Clinical_vs_GSVA <- anova(model_clinical, model_GSVA, test = "Chisq") 
Alliance_Clinical_vs_GSVA # p-value = 0.0528

```
## Incremental model building (stepwise covariate addition)

```{r incremental_model, echo=FALSE}
# Named list — goes directly into the Model column of the output table


predictor_sets <- list(
  "Sex + Smoking"          = c("gender", "smoking"),
  "+ ACQ-6"                = c("gender", "smoking", "acq6_score"),
  "+ FEV1"                 = c("gender", "smoking", "acq6_score", "B_FEV1PNVG"),
  "+ Exacerbation history" = c("gender", "smoking", "acq6_score", "B_FEV1PNVG", "ex_last_y"),
  "+ Blood eosinophils"    = c("gender", "smoking", "acq6_score", "B_FEV1PNVG", "ex_last_y", "LABEOSV"),
  "+ GINA"                 = c("gender", "smoking", "acq6_score", "B_FEV1PNVG", "ex_last_y", "LABEOSV", "GINA_mild_severe"),
  "+ GSVA"                 = c("gender", "smoking", "acq6_score", "B_FEV1PNVG", "ex_last_y", "LABEOSV", "GINA_mild_severe", "exacerbation_genes")
)

```

## Cross-validation: repeated 5-fold CV (500 repeats)

This section uses the functions defined in `cv_functions.R`.

```{r cross_validation, echo=FALSE, cache=TRUE}

# Set up caret trainControl 
train_ctrl <- 
  trainControl(
    method          = "repeatedcv",
    number          = 5,
    repeats         = 500,
    classProbs      = TRUE,
    summaryFunction = twoClassSummary,
    savePredictions = "final"
  )

```

## Run incremental models with Cross-validation 

```{r incremental_cv_train, echo=FALSE, cache=TRUE}
# All 7 models in one parallel batch — cached after first knit
incremental_cv_models <- run_incremental_cv_models(
  predictor_sets = predictor_sets,
  data           = model_table,
  train_ctrl     = train_ctrl,
  seed           = 42
)
```

## Create a table with all the metrics from incremental models

```{r incremental_cv_table, echo=FALSE}

incremental_table <- build_incremental_cv_table(
  predictor_sets = predictor_sets,
  cv_models      = incremental_cv_models
)

print(incremental_table, n = Inf)
write.csv(incremental_table, OUTPUTS$alliance_incremental_table)

```


## Add two genes to the model


```{r gsva_model, echo=FALSE}
model_twogenes = glm(exacerbation ~ gender + smoking + acq6_score +  B_FEV1PNVG + ex_last_y + LABEOSV + GINA_mild_severe + TNFAIP6 + ENSG00000287937, data = model_table, family = binomial)
summary(model_twogenes)

coef_table <- summary(model_twogenes)$coefficients %>%
  as.data.frame() %>%
  tibble::rownames_to_column("term") %>%
  as_tibble() %>%
  mutate(across(where(is.numeric), ~ round(.x, 3)))
                              
write.csv(coef_table, OUTPUTS$alliance_2genes_model)

```

