---
title: "ATLANTIS_exacerbations"
author: "Tatiana Karp, Rui Mercalo"
date: "2025-07-15"
output: html_document
---

```{r setup, include=FALSE}

library(dplyr)
library(edgeR)
library(ggplot2)
library(GSVA)
library(pROC)
library(biomaRt)
library(ggrepel)
library(stringr)
library(here)

source(here("config.R"))
knitr::opts_knit$set(root.dir = ROOT_DIR)

# Source the cross-validation helper functions
source(here("R/cv_functions.R"))

```

## Prepare the metadata

```{r metadata, echo=FALSE}
atlantis_exacerbations <- read.csv(PATHS$metadata) %>%
  dplyr::select(-X) %>%
  filter(QC_check == 'YES') %>%
  mutate(
    gender                = as.factor(gender),
    smoking.status        = as.factor(smoking.status),
    exacerbation          = as.factor(exacerbation),
    exacerbation_f = factor(if_else(exacerbation == TRUE, "Yes", "No"), levels = c("Yes", "No")),
    time_to_exacerbation  = as.numeric(time_to_exacerbation),
    follow_up             = as.numeric(follow_up),
    Ex_freq               = as.factor(if_else(NUM_EX_D > 1, 'freq', 'not_freq'))
  )

# Filter out samples with follow-up < 250 days
atlantis_exacerbations_decreased <- atlantis_exacerbations %>%
  filter(follow_up >= 250)
```

## Add expression data and perform differential gene expression analysis

```{r DGE, echo=FALSE}
expression.data <- read.csv(PATHS$expression, header = TRUE) %>%
  tibble::column_to_rownames("Gene") %>%
  dplyr::select(all_of(atlantis_exacerbations_decreased$GenomeScan_ID)) %>%
  as.matrix()

all(colnames(expression.data) == atlantis_exacerbations_decreased$GenomeScan_ID)

# DE: yes/no exacerbations
design <- model.matrix(
  ~0 + exacerbation + age + gender + smoking.status,
  data = atlantis_exacerbations_decreased
)

DGEL_exac <- edgeR::DGEList(expression.data)
keep <- edgeR::filterByExpr(DGEL_exac, design)
DGEL_exac  <- DGEL_exac[keep, , keep.lib.sizes = FALSE]
DGEL_exac  <- edgeR::calcNormFactors(DGEL_exac, method = "TMM")

TMM_expr <- cpm(DGEL_exac, normalized.lib.sizes = TRUE, log = TRUE)

DGEL_exac  <- edgeR::estimateDisp(DGEL_exac, design)
fit_exac   <- edgeR::glmQLFit(DGEL_exac, design, legacy = TRUE)
contrasts  <- limma::makeContrasts(
  EX_binary = exacerbationTRUE - exacerbationFALSE,
  levels    = design
)
qlf_exac <- edgeR::glmQLFTest(fit_exac, contrast = contrasts[, "EX_binary"])
summary(decideTests(qlf_exac))

de.results.exac <- edgeR::topTags(qlf_exac, n = nrow(DGEL_exac))$table %>%
  tibble::rownames_to_column("Gene")
```

## Add gene names via biomaRt

```{r gene_names, echo=FALSE}
ensembl38  <- useEnsembl(biomart = "genes", dataset = "hsapiens_gene_ensembl", version = 101)
dataprobes <- getBM(
  filters    = "ensembl_gene_id",
  attributes = c("ensembl_gene_id", "external_gene_name"),
  values     = de.results.exac$Gene,
  mart       = ensembl38
)

de.results.exac <- de.results.exac %>%
  dplyr::left_join(dataprobes, by = c("Gene" = "ensembl_gene_id"))
```

## Volcano plot

```{r volcano, echo=FALSE}
volcano <- ggplot(data = de.results.exac, aes(x = logFC, y = -log10(FDR))) +
  geom_point(aes(color = ifelse(
    (FDR < 0.05) & (logFC > 0), 'firebrick',
    ifelse((FDR < 0.05) & (logFC < 0), 'blue', 'gray')
  ))) +
  theme_classic() +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", linewidth = 0.5) +
  geom_text_repel(
    aes(label = ifelse(
      ((FDR < 0.05) & (abs(logFC) > 1)) | (FDR < 0.000001),
      ifelse(!is.na(external_gene_name), external_gene_name, ''), ''
    )),
    max.overlaps = 50, direction = 'both',
    alpha = 0.6, box.padding = 0.3, point.padding = 0.5
  ) +
  xlab(expression(log[2] ~ fold ~ change)) +
  ylab(expression(-log[10] ~ FDR)) +
  scale_color_identity(
    name   = '',
    breaks = c('firebrick', 'blue', 'gray'),
    labels = c('higher in exacerbators', 'lower in exacerbators', 'not significant'),
    guide  = "legend"
  ) +
  xlim(-2.5, 2.5) +
  theme(
    axis.text        = element_text(size = 15),
    axis.title       = element_text(size = 15, face = "plain"),
    legend.position  = "bottom",
    panel.grid.major = element_line(colour = "grey70", size = 0.1),
    panel.grid.minor = element_blank(),
    panel.background = element_blank(),
    axis.line        = element_line(colour = "black", linewidth = 0.6),
    legend.text      = element_text(size = 12),
    legend.key       = element_rect(fill = "white"),
    plot.title       = element_text(hjust = 0.5, size = 15)
  )

saveRDS(volcano, OUTPUTS$volcano_rds)
png(OUTPUTS$volcano_png, width = 1100, height = 1200, res = 150)
print(volcano)
dev.off()
```

## Calculate ssGSEA score for DE genes

```{r ssGSEA, echo=FALSE}
geneSet   <- list(exacerbation_genes = de.results.exac[de.results.exac$FDR < 0.05, 'Gene'])
ssgseapar <- ssgseaParam(as.matrix(TMM_expr), geneSet, normalize = FALSE)

gsva_score <- gsva(ssgseapar, verbose = TRUE) %>%
  t() %>%
  as.data.frame() %>%
  tibble::rownames_to_column('Sample')

atlantis_exacerbations_decreased <- atlantis_exacerbations_decreased %>%
  left_join(gsva_score, by = c("GenomeScan_ID" = "Sample")) %>%
  rename("gsva_score_exacerbation_genes" = "exacerbation_genes")
```

## Prepare clinical data for exacerbation prediction

```{r prepare_clinical_model, echo=FALSE}
clinical_table <- clinical_table <- read.csv(PATHS$clinical, header = TRUE, na.strings = c("", "NA")) %>%
  dplyr::select(c('PT', 'GINA', 'LABEOSV', 'B_FEV1PNVG', 'acq6_score', 'NUM_EX'))
clinical_table <- clinical_table[!duplicated(clinical_table$PT), ] %>%
  mutate(GINA = as.factor(GINA))

model_table <- atlantis_exacerbations_decreased %>%
  left_join(clinical_table, by = 'PT') %>%
  dplyr::select(gender, smoking.status, GenomeScan_ID, GINA, LABEOSV,
                B_FEV1PNVG, Ex_freq, exacerbation, exacerbation_f, follow_up,
                NUM_EX_D, NUM_EX, acq6_score, gsva_score_exacerbation_genes) %>%
  mutate(
    ex_last_y      = as.factor(if_else(NUM_EX > 0, 'YES', 'NO')),
    GINA_mild_severe = if_else(GINA %in% c('1','2','3'), 'mild', 'severe'),
    smoking        = if_else(smoking.status %in% c('Non.smoker'), 'never', 'ex_cur_smoker')
  ) %>%
  filter(!is.na(LABEOSV), !is.na(B_FEV1PNVG))

table(model_table$exacerbation)
```

## Logistic regression: in-sample model fit

```{r clinical_model, echo=FALSE}
model_clinical <- glm(
  exacerbation ~ gender + smoking + acq6_score + B_FEV1PNVG +
    ex_last_y + LABEOSV + GINA_mild_severe,
  data   = model_table,
  family = binomial
)
summary(model_clinical)

prob_clinical <- stats::predict(model_clinical, type = "response")
roc_clinical  <- roc(model_table$exacerbation ~ prob_clinical,
                     plot = TRUE, print.auc = TRUE)
auc(roc_clinical)
```

```{r gsva_model, echo=FALSE}
model_GSVA <- glm(
  exacerbation ~ gender + smoking + acq6_score + B_FEV1PNVG +
    ex_last_y + LABEOSV + GINA_mild_severe + gsva_score_exacerbation_genes,
  data   = model_table,
  family = binomial
)
summary(model_GSVA)

prob_GSVA <- stats::predict(model_GSVA, type = "response")
roc_GSVA  <- roc(model_table$exacerbation ~ prob_GSVA,
                 plot = TRUE, print.auc = TRUE)
auc(roc_GSVA)
```

## Compare clinical vs GSVA model (likelihood-ratio test)

```{r compare_model, echo=FALSE}
ATLANTIS_Clinical_vs_GSVA <- anova(model_clinical, model_GSVA, test = "Chisq")
ATLANTIS_Clinical_vs_GSVA
```

## ROC curves

```{r roc, echo=FALSE}
ATLANTIS_roc_clinical <- roc(model_table$exacerbation, prob_clinical)
ATLANTIS_roc_GSVA     <- roc(model_table$exacerbation, prob_GSVA)

roc_df <- data.frame(
  fpr   = c(1 - ATLANTIS_roc_clinical$specificities,
            1 - ATLANTIS_roc_GSVA$specificities),
  tpr   = c(ATLANTIS_roc_clinical$sensitivities,
            ATLANTIS_roc_GSVA$sensitivities),
  model = rep(c("Clinical", "Clinical + GSVA"),
              each = length(ATLANTIS_roc_clinical$specificities))
)

ggplot(roc_df, aes(x = fpr, y = tpr, color = model)) +
  geom_line(linewidth = 1.2) +
  geom_abline(linetype = "dashed", color = "gray") +
  labs(
    title = "ROC Curve: Clinical vs Clinical + GSVA (ATLANTIS)",
    x     = "False Positive Rate (1 - Specificity)",
    y     = "True Positive Rate (Sensitivity)"
  ) +
  theme_minimal() +
  scale_color_manual(values = c("blue", "red"))
```

## Incremental model building (stepwise covariate addition)

```{r incremental_model, echo=FALSE}
# Named list — goes directly into the Model column of the output table


predictor_sets <- list(
  "Sex + Smoking"          = c("gender", "smoking"),
  "+ ACQ-6"                = c("gender", "smoking", "acq6_score"),
  "+ FEV1"                 = c("gender", "smoking", "acq6_score", "B_FEV1PNVG"),
  "+ Exacerbation history" = c("gender", "smoking", "acq6_score", "B_FEV1PNVG", "ex_last_y"),
  "+ Blood eosinophils"    = c("gender", "smoking", "acq6_score", "B_FEV1PNVG", "ex_last_y", "LABEOSV"),
  "+ GINA"                 = c("gender", "smoking", "acq6_score", "B_FEV1PNVG", "ex_last_y", "LABEOSV", "GINA_mild_severe"),
  "+ GSVA"                 = c("gender", "smoking", "acq6_score", "B_FEV1PNVG", "ex_last_y", "LABEOSV", "GINA_mild_severe", "gsva_score_exacerbation_genes")
)

```

## Cross-validation: repeated 5-fold CV (500 repeats)

This section uses the functions defined in `cv_functions.R`.

```{r cross_validation, echo=FALSE, cache=TRUE}

# Set up caret trainControl 
train_ctrl <- 
  trainControl(
    method          = "repeatedcv",
    number          = 5,
    repeats         = 500,
    classProbs      = TRUE,
    summaryFunction = twoClassSummary,
    savePredictions = "final"
  )

```

## Run incremental models with Cross-validation 

```{r incremental_cv_train, echo=FALSE, cache=TRUE}
# All 7 models in one parallel batch — cached after first knit
incremental_cv_models <- run_incremental_cv_models(
  predictor_sets = predictor_sets,
  data           = model_table,
  train_ctrl     = train_ctrl,
  seed           = 42
)
```

## Create a table with all the metrics from incremental models

```{r incremental_cv_table, echo=FALSE}

incremental_table <- build_incremental_cv_table(
  predictor_sets = predictor_sets,
  cv_models      = incremental_cv_models
)

print(incremental_table, n = Inf)

```

## Plot sensitivity by cumulative covariate addition

```{r sensitivity_plot, echo=FALSE}
model_perf <- incremental_table %>%
  dplyr::select(Model, Sensitivity, Sensitivity_CI_low, Sensitivity_CI_high) %>%
  mutate(
    Model               = factor(Model, levels = Model),
    Sensitivity         = Sensitivity * 100,
    Sensitivity_CI_low  = Sensitivity_CI_low * 100,
    Sensitivity_CI_high = Sensitivity_CI_high * 100
  )

custom_labels <- c(
  "Sex + Smoking", "+ ACQ-6",
  expression("+ FEV"[1] * " % pred"),
  "+ Exacerbation history", "+ Blood eosinophils", "+ GINA", "+ GSVA"
)

ggplot(model_perf, aes(x = Model, y = Sensitivity)) +
  geom_bar(stat = "identity", fill = "#2c7fb8", alpha = 0.85) +
  geom_errorbar(
    aes(ymin = Sensitivity_CI_low, ymax = Sensitivity_CI_high),
    width  = 0.25,
    colour = "gray30",
    linewidth = 0.7
  ) +
  geom_text(aes(label = paste0(Sensitivity, "%")),
            vjust = -0.5, size = 4) +
  labs(
    title = "CV Sensitivity (True Positive Rate) with each additional variable",
    x     = "Model (Cumulative Addition of Predictors)",
    y     = "Sensitivity (%)"
  ) +
  theme_minimal(base_size = 13) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_x_discrete(labels = custom_labels) +
  ylim(0, 50)
```
